---
title: "Class 1 - Intro to R"
author: "Connor Guest"
date: "6/7/2019"
output: 
html_document:
    theme: cosmo
    highlight: monochrome
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install packages, echo = TRUE, message = FALSE, warning = FALSE}
# install.packages(c("stringr", "devtools", "dplyr", "tidyr", "ggplot2", "gganimate", "gifski", "png", "lubridate", "readr", "wesanderson"))
```


```{r load packages, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
library(gganimate)
library(devtools)
library(stringr)
library(tidyr)
library(gifski)
library(png)
library(wesanderson)

theme_set(
  theme_minimal() +
    theme(legend.position = "right")
  )
```


## Importing Data 

```{r import, }
failed_trans <- read_csv("Branch Failed Transactions for April.csv")
```

```{r inspect,  echo = TRUE, results = 'hide'}
# View dimensions of failed_trans
dim(failed_trans)

# Inspect first 6 rows of failed_trans
head(failed_trans)

# View column names of failed_trans
names(failed_trans)

# Look at structure of sales
str(failed_trans)

# View a summary of sales
summary(failed_trans)

# Get a glimpse of sales
glimpse(failed_trans)
```

## Cleaning Data

```{r,  echo = TRUE, results = 'hide'}
# rename columns 
failed_trans_1 <- rename(failed_trans, datetime = "Transaction Date", amount = "Transaction amount", error_category = errorcategory)

# View column names of failed_trans
names(failed_trans_1)
```

Transaction over time 
```{r Failures over time,  echo = TRUE, results = 'hide'}
# you can use substring or change the format

failed_trans_1$date <- as.Date(failed_trans_1$datetime, "%Y/%m/%d")
failed_trans_1$hour <- hour(failed_trans_1$datetime)
failed_trans_1$weekday <- wday(failed_trans_1$datetime)

str(failed_trans_1)

```

We now have an understanding of the structure of the dataset. Almost 200,000 rows and 14 variables.

Let's start answering some basic questions.

Question 1 - Are there any patterns in the number of transactions over the month? 

```{r}
failed_trans_by_day <- failed_trans_1 %>% 
  group_by(date) %>% 
  summarize(count = n())

line_chart <- failed_trans_by_day %>%
  ggplot(aes(x = date , y = count)) +
  geom_line(color = "#3B9AB2") +
  labs( x = "Date", y = "Number of Failed Transaction",
        title ="Failed Transactions Over Time", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  expand_limits(y = 0)  +
  theme_minimal()

# line chart using gganimate
line_chart + 
  geom_point(aes(group = seq_along(date))) +
  transition_reveal(date)

# line_chart + geom_point()

```


Are there any patterns by hour of the day?



```{r}
failed_trans_by_hour <- failed_trans_1 %>%
  group_by(hour) %>%
  filter(hour != "7") %>%
  summarize(count = n())

bar_chart_by_hour <- failed_trans_by_hour %>%
  ggplot(aes(x = hour , y = count)) +
  geom_col(fill = "#3B9AB2") +
  labs( x = "Hour of the Day", y = "Number of Failed Transaction",
        title ="Failed Transactions Over A Day on Avg", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  expand_limits(y = 0)  +
  theme_minimal()

bar_chart_by_hour

# bar chart using gganimate
# bar_chart_by_hour + 
#   transition_states(hour, wrap = FALSE) +
#   shadow_mark() +
#   enter_grow() +
#   enter_fade()

failed_trans_by_wday <- failed_trans_1 %>%
  group_by(weekday) %>%
  summarize(count = n())

bar_chart_by_wday <- failed_trans_by_wday %>%
  ggplot(aes(x = weekday, y = count)) +
  geom_col(fill = "#3B9AB2") +
  labs( x = "Day of the Week", y = "Number of Failed Transaction",
        title ="Failed Transactions by Day of Week", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  expand_limits(y = 0)  +
  theme_minimal()

bar_chart_by_wday

```



```{r counting}
bank_fail <- failed_trans_1 %>% 
  group_by(bank) %>%
  summarize(count = n()) %>%
  filter(!is.na(bank)) %>%
  arrange(desc(count)) %>% head(22)

bank_fail %>%
  ggplot(aes(x = reorder(bank, count) , y = count)) +
  geom_col(fill = "#3B9AB2") + coord_flip() +
  labs( x = "Banks", y = "Number of Failed Transaction",
        title ="Comparing Failed Transactions by Bank", 
        subtitle = "the top 22 Banks out of 44", 
        caption = "All data from month of April" )

```

Which individual had the most failed transactions? What can we learn from them?
```{r}
whos_failing <- failed_trans_1 %>%
  group_by(email) %>%
  mutate(fail_num = n()) %>%
  arrange(desc(fail_num)) %>%
  top_n(50)
```


```{r}
error_category_fail <- failed_trans_1 %>% 
  group_by(error_category) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))

error_category_fail %>%
  ggplot(aes(x = reorder(error_category, count) , y = count)) +
  geom_col(fill = "#3B9AB2") + coord_flip() +
  labs( x = "Error Category", y = "Number of Failed Transactions",
        title ="Comparing Failed Transactions by Error Category", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  theme_minimal()
```
```{r}
bank_error_category_fail <- failed_trans_1 %>% 
  filter(error_category == "bank") %>% 
  group_by(gateway_response) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))

bank_error_category_fail %>%
  ggplot(aes(x = reorder(gateway_response, count) , y = count)) +
  geom_col(fill = "#3B9AB2") + coord_flip() +
  labs( x = "Response", y = "Number of Failed Transactions",
        title ="Comparing Failed Transactions by Gateway Response", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  theme_minimal()
```

```{r}
individual_bank_error_category_fail <- failed_trans_1 %>% 
  filter(error_category == "bank") %>%
  filter(bank == c('Guaranty Trust Bank', 'Access Bank','Access Bank', 'First Bank of Nigeria', 'Zenith Bank', 'United Bank for Africa', 'Access Bank (Diamond)'))  %>%
  group_by(bank, gateway_response) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))

individual_bank_error_category_fail %>%
  ggplot(aes(x = reorder(gateway_response, count) , y = count)) +
  geom_col(fill = "#3B9AB2") + coord_flip() +
  labs( x = "Response", y = "Number of Failed Transactions",
        title ="Comparing Failed Transactions by Gateway Response", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  theme_minimal() +
  facet_wrap(~bank)
```
 
 
```{r}
gateway_fail <- failed_trans_1 %>% 
  group_by(gateway) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))

gateway_fail %>%
  ggplot(aes(x = reorder(gateway, count) , y = count)) +
  geom_col(fill = "#3B9AB2") + coord_flip() +
  labs( x = "Banks", y = "Number of Failed Transaction",
        title ="Comparing Failed Transactions by Gateway", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  theme_minimal()

```

We see that there are 0 failed 737 transactions. This means that they are being considered abandoned? 

```{r}
# Facetting
bank_gateway_fail <- failed_trans_1 %>% 
  filter(bank == c('Guaranty Trust Bank', 'Access Bank','Access Bank', 'First Bank of Nigeria', 'Zenith Bank', 'United Bank for Africa', 'Access Bank (Diamond)'))  %>%
  group_by(bank, gateway) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))

bank_gateway_fail %>%
  ggplot(aes(x = reorder(gateway, count) , y = count)) +
  geom_col(fill = "#3B9AB2") + coord_flip() +
  labs( x = "Banks", y = "Number of Failed Transaction",
        title ="Comparing Failed Transactions by Gateway", 
        subtitle = "", 
        caption = "All data from month of April" ) +
  theme_minimal() + facet_wrap(~bank)
```
