---
title: "Data Structure Fundamentals"
author: "Connor Guest"
date: "6/7/2019"
output: 
html_document:
    theme: cosmo
    highlight: monochrome
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Data Fundamentals
The following is an overview of some of the fundamentals we have not yet covered including: 
Matrices
Vectors
Variables -  Categorical and Continuous
Factors, Levels - converting categorical variables to factors
Data Frames
Lists

Go through and try to understnad what each line is doing. Do some research on your own to further lear nabout the different topics if something is not making sense :) 

## Matrices
In R, a matrix is a collection of elements of the same data type (numeric, character, or logical) arranged into a fixed number of rows and columns. Since you are only working with rows and columns, a matrix is called two-dimensional.

```{r matrices}
# Construct a matrix with 3 rows that contain the numbers 1 up to 9
matrix(1:9, byrow = TRUE, nrow = 3)
```

```{r star wars}
# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Construct matrix
star_wars_matrix <- matrix(c(new_hope, empire_strikes, return_jedi), nrow = 3, byrow = TRUE)
star_wars_matrix

# Vectors region and titles, used for naming
region <- c("US", "non-US")
titles <- c("A New Hope", "The Empire Strikes Back", "Return of the Jedi")

# Name the columns with region
colnames(star_wars_matrix) <- region

# Name the rows with titles
rownames(star_wars_matrix) <- titles

# Print out star_wars_matrix
star_wars_matrix

# Calculate worldwide box office figures
worldwide_vector <- rowSums(star_wars_matrix)
worldwide_vector

# Bind the new variable worldwide_vector as a column to star_wars_matrix
all_wars_matrix <- cbind(star_wars_matrix, worldwide_vector)
all_wars_matrix
```


## Factors
The term factor refers to a statistical data type used to store categorical variables. The difference between a categorical variable and a continuous variable is that a categorical variable can belong to a limited number of categories. A continuous variable, on the other hand, can correspond to an infinite number of values.

It is important that R knows whether it is dealing with a continuous or a categorical variable, as the statistical models you will develop in the future treat both types differently.

A good example of a categorical variable is sex. In many circumstances you can limit the sex categories to "Male", "Female", and "Gender Diverse". 

What would be another type of categorical variable? Remember it is a variable with a limited number of categories. (ex. Bank, Region, Company, Error Response, Channel, Zipcode)

What would be a quantitative variable? Remember it is a variable with an unlimited number of categories. (ex. transaction amount, GPA, file size)

The different categories of a categorical variable are called "Levels" as you will see below.
```{r}
# Sex vector
sex_vector <- c("Male", "Female", "Female", "Male", "Male")

# Convert sex_vector to a factor
factor_sex_vector <- factor(sex_vector)

# Print out factor_sex_vector
factor_sex_vector

# Code to build factor_survey_vector
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)

factor_survey_vector

# Specify the levels of factor_survey_vector
levels(factor_survey_vector) <- c("Female", "Male")

factor_survey_vector

# Generate summary for survey_vector
# Note that the class is still a character
summary(survey_vector)

# Generate summary for factor_survey_vector
summary(factor_survey_vector)
```

```{r}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
speed_vector

# Convert speed_vector to ordered factor vector
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast")) # you are specifying the order here

# Print factor_speed_vector
factor_speed_vector
summary(factor_speed_vector)
```

## Data Frames
In a data frame, variables are the columns and inputs are the rows. 

Applying the str() function (try it below) will often be the first thing that you do when receiving a new data set or data frame. It is a great way to get more insight in your data set before diving into the real analysis.  

For the mtcars dataframe, it tells you:  

The total number of observations (e.g. 32 car types)  
The total number of variables (e.g. 11 car features)  
A full list of the variables names (e.g. mpg, cyl ... )  
The data type of each variable (e.g. num)  
The first observations of each variable

mtcars is a dataframe that is available to everyone in R without importing it. You will see it come up frequently during your your data science learning journey.

```{r}
mtcars <- mtcars
head(mtcars)
tail(mtcars)
# str() tells you the structure 
str(mtcars)
```

Subsetting dataframes
```{r planets}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune")
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE) 

# Create a data frame from the vectors using data.frame()
planets_df <- data.frame(name, type, diameter, rotation, rings)

rings_vector <- planets_df$rings
rings_vector

# Select all columns from the planets that have rings (rings = TRUE)
planets_df[rings_vector,]

#Subset - same output as above, just a different method
subset(planets_df, subset = rings)

subset(planets_df, subset = diameter < 1)

# Use order() to create positions
positions <- order(planets_df$diameter) 
positions

# Use positions to sort planets_df
planets_df[positions,]
```

## Lists
A list in R allows you to gather a variety of objects under one name (that is, the name of the list) in an ordered way. These objects can be matrices, vectors, data frames, even other lists, etc. It is not even required that these objects are related to each other in any way.

You could say that a list is some kind super data type: you can store practically any piece of information in it!

Just like on your to-do list, you want to avoid not knowing or remembering what the components of your list stand for. That is why you should give names to them:
```{r}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 

# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)

# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct list with these different elements:
my_list <- list(my_vector, my_matrix, my_df)


names(my_list) <- c("vec", "mat", "df")
# you can add to lists by using the c() command

my_list_full <- c(my_list, year = 1980)

str(my_list_full)
```
