---
title: "Class 2 - data frames, dplyr, ggplot2"
author: "Connor Guest"
date: "6/7/2019"
output: 
html_document:
    theme: cosmo
    highlight: monochrome
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Class 2 - dplyr & ggplot2

## dplyr overview
dplyr is a package which provides a set of tools for efficiently manipulating datasets in R.  

dplyr is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges:  
  
* mutate() adds new variables that are functions of existing variables  
* select() picks variables based on their names.  
* filter() picks cases based on their values.  
* summarise() reduces multiple values down to a single summary.  
* arrange() changes the ordering of the rows. 

[see dplyr documentation](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8)

## ggplot2 overview
ggplot2 is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.

It’s hard to succinctly describe how ggplot2 works because it embodies a deep philosophy of visualisation. However, in most cases you start with ggplot(), supply a dataset and aesthetic mapping (with aes()). You then add on layers (like geom_point() or geom_histogram()), scales (like scale_colour_brewer()), faceting specifications (like facet_wrap()) and coordinate systems (like coord_flip()).

[see ggplot2 documentation](https://ggplot2.tidyverse.org/)
```{r install packages, include=false}
# Uncomment the line below and make sure to install the tidyverse - a collection of the most important R 

install.packages("tidyverse")
install.packages("gapminder")
install.packages("hflights")
```

## Load Libraries
```{r load libraries, echo=false}
library(tidyverse)
```

## Playing with the Gap Minder Data Set using dplyr
We will learn how to do similar data visualizations to the legendary Hans Rosling. [Watch Video Here](https://www.youtube.com/watch?v=jbkSRLYSojo)

```{r}
library(gapminder) # loads the gapminder dataset

# load the data frame into your global environment
gapminder <- tbl_df(gapminder)

# make sure to note: how many variables there are, the variable names, the data type, the number of observations. 
# Use one of the commands we used last class.

glimpse(gapminder)

``` 

```{r}
#use the filter function and a pipe operator to get only data form the year 1957
gapminder_1957_or_africa <- gapminder %>% filter(year == 1957 | continent == "Africa")


gapminder_africa <- gapminder %>% filter(continent == c("Africa", "Asia", "America"))



#Using arrange()
# Sort in ascending order of lifeExp
gapminder %>% arrange(lifeExp)


arrange(gapminder, lifeExp)
  
# Sort in descending order of lifeExp
gapminder %>% arrange(desc(lifeExp))

# Filter for the year 1957, then arrange in descending order of population
gapminder %>% filter(year == 1957) %>% arrange(desc(pop))



# Using the mutate() command
# Use mutate to change lifeExp to be in months
gap1 <- gapminder %>% mutate(lifeExp_month = lifeExp * 12)

gap1 <- gap1 %>% select(-lifeExp_month)


# Use mutate to create a new column called lifeExpMonths
gapminder %>% mutate(lifeExpMonths = lifeExp * 12)


# Filter(year 2007), mutate(create new column lifeExpMonths), and arrange the gapminder dataset to show the country with the highest life exp
gapminder %>% 
filter(year == 2007) %>% 
mutate(lifeExpMonths = lifeExp * 12) %>% 
arrange(desc(lifeExpMonths))


# Create a new dataframe from year 1952, create a new column for total gdp, and arrange so that the richest country is at the top
GDP_1952 <- gapminder %>% 
filter(year == 1952) %>% 
mutate(total_gdp = gdpPercap * pop) %>% 
arrange(desc(total_gdp))


GDP_1952$total_gdp <- format(GDP_1952$total_gdp, scientific=F)

format(xx, scientific=F)
 

```


## ggplot2

```{r}
library(ggplot2)
gapminder_1952 <- gapminder %>% filter(year == 1952)

gapminder_1952 %>% 
  ggplot(aes(x=lifeExp, y=gdpPercap)) + geom_point()



gapminder_1952 %>% filter(gdpPercap > 90000)


#population vs life expectancy
gapminder_1952 %>% 
  ggplot(aes(x = lifeExp, y = pop)) + 
  geom_point()

# Log 10 Scale
# Change this plot to put the x-axis on a log scale
gapminder_1952 %>% 
  ggplot(aes(x = pop, y = lifeExp)) +
  geom_point()+
  scale_x_log10()

# add color for continent and size for gdpPercap
gapminder_1952 %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  geom_point()+
  scale_x_log10()

# facetting
gapminder_1952 %>% 
  ggplot(aes(x = pop, y = lifeExp, color = continent, size = gdpPercap)) +
  geom_point()+
  scale_x_log10() + 
  facet_wrap(~ continent)

gapminder %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  geom_point()+
  scale_x_log10() + 
  facet_wrap(~ year)
```

## Summarize

```{r}
# Summarize to find the median life expectancy
gapminder %>% 
summarize(medianLifeExp = median(lifeExp))

# Filter for 1957 then summarize the median life expectancy and the maximum GDP per capita
gapminder %>% 
filter(year == 1957) %>%
summarize(medianLifeExp = median(lifeExp), 
          maxGdpPercap = max(gdpPercap))

gapminder %>% 
group_by(year) %>%
summarize(medianLifeExp = median(lifeExp), 
          maxGdpPercap = max(gdpPercap))

# Find median life expectancy and maximum GDP per capita in each continent/year combination

gapminder %>% 
group_by(continent, year) %>%
summarize(medianLifeExp = median(lifeExp), 
          maxGdpPercap = max(gdpPercap))

#Visualize with a scatter plot
by_year_continent <- gapminder %>%
  group_by(year, continent) %>%
  summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap),
            totalPop = sum(pop))

# Create a scatter plot showing the change in medianLifeExp over time
by_year_continent %>% 
ggplot(aes(x = year, y = medianLifeExp, color = continent, size = totalPop)) +
geom_point() +
expand_limits(y = 0)
```


```{r}
# Summarize medianGdpPercap within each continent within each year: by_year_continent
by_year_continent1 <- gapminder %>%
  group_by(continent, year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Plot the change in medianGdpPercap in each continent over time
by_year_continent1 %>% 
ggplot(aes(x = year, y = medianGdpPercap, color = continent)) +
geom_point() +
expand_limits(y = 0)

# look into Nigeria
Africadata <- gapminder %>% 
  filter(country == c("Nigeria", "Ghana", "Egypt")) %>%
  group_by(country, year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

Africadata %>% 
  ggplot(aes(x = year, y = medianGdpPercap, color = country)) +
  geom_point() + 
  expand_limits(y = 0)

# Summarize the median GDP and median life expectancy per continent in 2007
by_continent_2007 <- gapminder %>%
filter(year == 2007) %>%
group_by(continent) %>%
summarize(medianLifeExp = median(lifeExp),
          medianGdpPercap = median(gdpPercap),
          total_population = sum((pop)))

# Use a scatter plot to compare the median GDP and median life expectancy
by_continent_2007 %>%
ggplot(aes(x = medianGdpPercap, y = medianLifeExp, color = continent, size = total_population)) + 
geom_point() +
expand_limits(y = 0)  
```

## Line, bar, histogram, and box plots

```{r}
# Line Plot

# Summarize the median gdpPercap by year, then save it as by_year
by_year <- gapminder %>%
group_by(year) %>%
summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap over time
by_year %>%
ggplot(aes(x = year, y = medianGdpPercap)) +
geom_line() +
expand_limits(y=0)
```

```{r}
# Summarize the median gdpPercap by year & continent, save as by_year_continent
by_year_continent <- gapminder %>%
group_by(continent, year) %>%
summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap by continent over time
by_year_continent %>%
ggplot(aes(x = year, y = medianGdpPercap, color = continent)) +
geom_line() +
expand_limits(y=0)
```

Bar Plot
```{r}
# Bar Plot

# Summarize the median gdpPercap by year and continent in 1952
by_continent <- gapminder %>%
filter(year == 1952) %>%
group_by(continent) %>%
summarize(medianGdpPercap = median(gdpPercap))

# Create a bar plot showing medianGdp by continent
by_continent %>%
ggplot(aes(x = continent, y = medianGdpPercap)) +
geom_col() 
```

Multiple Filters
```{r}
# multiple filters 
# Filter for observations in the Oceania continent in 1952
oceania_1952 <- gapminder %>%
filter(continent == "Oceania" & year == 1952)

# Create a bar plot of gdpPercap by country
oceania_1952 %>%
ggplot(aes(x = country, y = gdpPercap)) +
geom_col() 
```

Histograms
```{r}
# Histogram - A histogram is useful for examining the distribution of a numeric variable. 
gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Create a histogram of population (pop), with x on a log scale
gapminder_1952 %>% ggplot(aes(x = pop)) + 
scale_x_log10() +
geom_histogram()
```

Box Plots
```{r}
#boxplots

gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Create a boxplot comparing gdpPercap among continents
gapminder_1952 %>%
ggplot(aes(x = continent, y = gdpPercap)) +
geom_boxplot() + 
scale_y_log10() + 
ggtitle("Comparing GDP per capita across continents")
```


Check out the [Gapminder Homepage](https://www.gapminder.org/) for more insightful graphics!


## More DPLYR Practice
```{r}
# install.packages("hflights")
library(hflights)
```

```{r}
head(hflights)
str(hflights)
summary(hflights)
dim(hflights)
names(hflights) 
```

```{r}
hflights <- tbl_df(hflights)
hflights
glimpse(hflights)

carriers <- hflights$UniqueCarrier
```

A bit of cleaning would be a good idea since the UniqueCarrier variable of hflights uses a confusing code system.

To do this, let's work with a lookup table, that comes in the form of a named vector. When you subset the lookup table with a character string (like the character strings in UniqueCarrier), R will return the values of the lookup table that correspond to the names in the character string. To see how this works, run following code in the console:

```{r}
two <- c("AA", "AS")
lut <- c("AA" = "American", 
         "AS" = "Alaska", 
         "B6" = "JetBlue")
two <- lut[two]
two
```

```{r using a lookup table}
# Both the dplyr and hflights packages are loaded into workspace
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")

# Add the Carrier column to hflights
hflights$Carrier <- lut[hflights$UniqueCarrier]

# Glimpse at hflights
glimpse(hflights)
```

Let's try a similar thing, but this time to change the labels in the CancellationCode column. This column lists reasons why a flight was cancelled using a non-informative alphabetical code. Execute

```{r}
unique(hflights$CancellationCode)

# The lookup table
lut <- c("A" = "carrier", "B" = "weather", "C" = "FFA", "D" = "security", "E" = "not cancelled")

# Add the Code column
hflights$Code <- lut[hflights$CancellationCode]

# Glimpse at hflights
glimpse(hflights)

unique(hflights$Code)
```

```{r}
# Print out a tbl with the four columns of hflights related to delay
select(hflights, ActualElapsedTime, AirTime, ArrDelay, DepDelay)

# Print out the columns Origin up to Cancelled of hflights
select(hflights, Origin:Cancelled)

# Answer to last question: be concise!
select(hflights, Year:DayOfWeek, ArrDelay:Diverted)
```

dplyr comes with a set of helper functions that can help you select groups of variables inside a select() call:

starts_with("X"): every name that starts with "X",
ends_with("X"): every name that ends with "X",
contains("X"): every name that contains "X",
matches("X"): every name that matches "X", where "X" can be a regular expression,
num_range("x", 1:5): the variables named x01, x02, x03, x04 and x05,
one_of(x): every name that appears in x, which should be a character vector.
Pay attention here: When you refer to columns directly inside select(), you don't use quotes. If you use the helper functions, you do use quotes.

```{r using helper functions}
# Print out a tbl containing just ArrDelay and DepDelay
select(hflights, ArrDelay, DepDelay)

# Print out a tbl as described in the second instruction, using both helper functions and variable names
select(hflights, UniqueCarrier, ends_with("Num"), starts_with("Cancel"))

# Print out a tbl as described in the third instruction, using only helper functions.
select(hflights, ends_with("time"), ends_with("Delay"))
```

mutate() creates new columns which are added to a copy of the dataset.

```{r}
# Add the new variable ActualGroundTime to a copy of hflights and save the result as g1.
g1 <- mutate(hflights, ActualGroundTime = ActualElapsedTime - AirTime)

# Add the new variable GroundTime to g1. Save the result as g2.
g2 <- mutate(g1, GroundTime = TaxiIn + TaxiOut)

# Add the new variable AverageSpeed to g2. Save the result as g3.
g3 <- mutate(g2, AverageSpeed = 60 * Distance / AirTime)

# Add a second variable loss_ratio to the dataset: m1
m1 <- mutate(hflights, loss = ArrDelay - DepDelay, loss_ratio = loss / DepDelay)

# Add the three variables as described in the third instruction: m2
m2 <- mutate(hflights, TotalTaxi = TaxiIn + TaxiOut, ActualGroundTime = ActualElapsedTime - AirTime, Diff = TotalTaxi - ActualGroundTime)
```

Looking at FILTER()
x < y, TRUE if x is less than y
x <= y, TRUE if x is less than or equal to y
x == y, TRUE if x equals y
x != y, TRUE if x does not equal y
x >= y, TRUE if x is greater than or equal to y
x > y, TRUE if x is greater than y
x %in% c(a, b, c), TRUE if x is in the vector c(a, b, c)

```{r}
# All flights that traveled 3000 miles or more
filter(hflights, Distance > 3000)

# All flights flown by one of JetBlue, Southwest, or Delta
filter(hflights, UniqueCarrier %in% c("JetBlue", "Southwest", "Delta"))

# All flights where taxiing took longer than flying
filter(hflights, TaxiIn + TaxiOut > AirTime)

```

```{r}
# These two are the same
# filter(df, a > 0 & b > 0)
# filter(df, a > 0, b > 0)

# All flights that departed before 5am or arrived after 10pm
filter(hflights, DepTime < 500 | ArrTime > 2200)

# All flights that departed late but arrived ahead of schedule
filter(hflights, DepDelay > 0, ArrDelay < 0)

# All flights that were cancelled after being delayed
filter(hflights, DepDelay > 0, Cancelled == 1)

# Select the flights that had JFK as their destination: c1
c1 <- filter(hflights, Dest == "JFK")

# Combine the Year, Month and DayofMonth variables to create a Date column: c2
c2 <- mutate(c1, Date = paste(Year, Month, DayofMonth, sep = "-"))

# Print out a selection of columns of c2
select(c2, Date, DepTime, ArrTime, TailNum)

# How many weekend flights flew a distance of more than 1000 miles but had a total taxiing time below 15 minutes?
filter(hflights, DayOfWeek %in% c(6,7), Distance > 1000, TaxiIn + TaxiOut < 15)
```

ARRANGE()
arrange() can be used to rearrange rows according to any type of data. If you pass arrange() a character variable, for example, R will rearrange the rows in alphabetical order according to values of the variable. If you pass a factor variable, R will rearrange the rows according to the order of the levels in your factor (running levels() on the variable reveals this order).
```{r}
# Definition of dtc
dtc <- filter(hflights, Cancelled == 1, !is.na(DepDelay))

# Arrange dtc by departure delays
arrange(dtc, DepDelay)

# Arrange dtc so that cancellation reasons are grouped
arrange(dtc, CancellationCode)

# Arrange dtc according to carrier and departure delays
arrange(dtc, UniqueCarrier, DepDelay)
```
By default, arrange() arranges the rows from smallest to largest. Rows with the smallest value of the variable will appear at the top of the data set. You can reverse this behavior with the desc() function. arrange() will reorder the rows from largest to smallest values of a variable if you wrap the variable name in desc() before passing it to arrange().

```{r}
# Arrange according to carrier and decreasing departure delays
arrange(hflights, UniqueCarrier, desc(DepDelay))

# Arrange flights by total delay (normal order).
arrange(hflights, DepDelay + ArrDelay)
```

summarize(), the last of the 5 verbs, follows the same syntax as mutate(), but the resulting dataset consists of a single row instead of an entire new column in the case of mutate().

```{r}
# Print out a summary with variables min_dist and max_dist
summarize(hflights, min_dist = min(Distance), max_dist = max(Distance))

# Print out a summary with variable max_div
summarize(filter(hflights, Diverted == 1), max_div = max(Distance))
```

```{r}
# Remove rows that have NA ArrDelay: temp1
temp1 <- filter(hflights, !is.na(ArrDelay))

# Generate summary about ArrDelay column of temp1
summarize(temp1, earliest = min(ArrDelay), average = mean(ArrDelay), latest = max(ArrDelay), sd = sd(ArrDelay))

# Keep rows that have no NA TaxiIn and no NA TaxiOut: temp2
temp2 <- filter(hflights, !is.na(TaxiIn), !is.na(TaxiOut))

# Print the maximum taxiing difference of temp2 with summarize()
summarize(temp2, max_taxi_diff = max(abs(TaxiIn - TaxiOut)))
```

dplyr provides several helpful aggregate functions of its own, in addition to the ones that are already defined in R. These include:

first(x) - The first element of vector x.
last(x) - The last element of vector x.
nth(x, n) - The nth element of vector x.
n() - The number of rows in the data.frame or group of observations that summarize() describes.
n_distinct(x) - The number of unique values in vector x.
Next to these dplyr-specific functions, you can also turn a logical test into an aggregating function with sum() or mean(). A logical test returns a vector of TRUE's and FALSE's. When you apply sum() or mean() to such a vector, R coerces each TRUE to a 1 and each FALSE to a 0. sum() then represents the total number of observations that passed the test; mean() represents the proportion.

```{r}
# Generate summarizing statistics for hflights
summarize(hflights,
          n_obs = n(),
          n_carrier = n_distinct(hflights$UniqueCarrier),
          n_dest = n_distinct(hflights$Dest))

# All American Airline flights
aa <- filter(hflights, UniqueCarrier == "American")

# Generate summarizing statistics for aa 
summarize(aa,
          n_flights = n(),
          n_canc = sum(Cancelled),
          avg_delay = mean(ArrDelay, na.rm = TRUE))
```

The %>% operator allows you to extract the first argument of a function from the arguments list and put it in front of it, thus solving the Dagwood sandwich problem.
```{r}
# Write the 'piped' version of the English sentences.
hflights %>%
mutate(diff = TaxiOut - TaxiIn) %>%
filter(!is.na(diff)) %>% 
summarize(avg = mean(diff))

hflights %>% 
mutate(RealTime = ActualElapsedTime + 100, mph = (60 * Distance) / RealTime) %>%
filter(!is.na(mph) & mph < 70) %>%
summarize(n_less = n(),
          n_dest = n_distinct(Dest),
          min_dist = min(Distance),
          max_dist = max(Distance))

hflights %>%
  mutate(
    RealTime = ActualElapsedTime + 100, 
    mph = 60 * Distance / RealTime
  ) %>%
  filter(mph < 105 | Cancelled == 1 | Diverted == 1) %>%
  summarize(n_non = n(),
            n_dest = n_distinct(Dest),
            min_dist = min(Distance),
            max_dist = max(Distance))

# Count the number of overnight flights
hflights %>%
filter(!is.na(DepTime), !is.na(ArrTime), DepTime > ArrTime) %>%
summarize(num = n())
```
 GROUP_BY
 
 group_by() lets you define groups within your data set. Its influence becomes clear when calling summarize() on a grouped dataset: summarizing statistics are calculated for the different groups separately.
 
```{r}
# Make an ordered per-carrier summary of hflights
hflights %>%
  group_by(UniqueCarrier) %>%
  summarize(
    p_canc = 100 * mean(Cancelled == 1),
    avg_delay = mean(ArrDelay, na.rm = TRUE)
  ) %>%
  arrange(avg_delay, p_canc)
```
 
 You can also combine group_by() with mutate(). When you mutate grouped data, mutate() will calculate the new variables independently for each group. This is particularly useful when mutate() uses the rank() function, that calculates within-group rankings. rank() takes a group of values and calculates the rank of each value within the group, e.g.
rank(c(21, 22, 24, 23))
has output
[1] 1 2 4 3
As with arrange(), rank() ranks values from the smallest to the largest.

```{r rank}
# Ordered overview of average arrival delays per carrier
hflights %>% 
filter(ArrDelay > 0, !is.na(ArrDelay)) %>%
group_by(UniqueCarrier) %>%
summarize(avg = mean(ArrDelay)) %>%
mutate(rank = rank(avg)) %>%
arrange(rank)
```

```{r}
# How many airplanes only flew to one destination?
hflights %>%
  group_by(TailNum) %>%
  summarize(ndest = n_distinct(Dest)) %>%
  filter(ndest == 1) %>%
  summarize(nplanes = n())

# Find the most visited destination for each carrier
hflights %>%
  group_by(UniqueCarrier, Dest) %>%
  summarize(n = n()) %>%
  mutate(rank = rank(desc(n))) %>%
  filter(rank == 1)
```











